Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2014-11-10 MS_revised.tex
fixing several typos

We contrasted these results with those obtained from analysing the data in
the software BAMM \cite{rabosky2014}. We found 
that the best model (best configuration) estimated 2 diversification shifts in
our MCC tree (posterior probability = 0.79) for the Ithomiini
 and Satyrini. Bayes factor (BF) values strongly 
favored this model in comparison with a constant rate model (no diversification
shifts, BF $>$ 1,884).

@@

Our BiSSE analysis, extended to take into account missing taxa and
phylogenetic uncertainty, shows a significantly higher net
diversification rate for Ithomiini taxa, which can be attributed to the
trait ``feeding on Solanaceae hostplants'' (Fig. 4). This is in
agreement with previoues findings using other statistical methods
\cite{fordyce2010}. Due to the fact that Ithomiini are virtually the only
nymphalids using Solanaceae as hostplants, it is possible that the trait
responsible for a higher diversification of Ithomiini might not be the
hostplant character. As noted by Maddison et al. (2007) \cite{maddison2007},
the responsible trait might be a codistributed character such as a trait
related to the ability to digest secondary metabolites.

@@

The increase in diversification inferred by MEDUSA ocurred after the
probable shift from Apocynaceae to Solanaceae, as the Solanaceae feeders
in the subtribes Melinaeina and Mechanitina are not included in the
diversification shift (shift number 5 in Fig. 1). The apparent
conflicting results from MEDUSA and BiSSE can be explained by the low
species-richness of the subtribes Melinaeina and Mechanitina compared to
the other subtribes included in the shift (52 versus 272 species). It
can be that MEDUSA is more conservative than BiSSE and is not including
Melinaeina and Mechanitina in the shift due to low species numbers.

@@

Our MultiMEDUSA approach to perform analyses on the posterior
distribution of trees found strong support for an increase in net
diversification rate in the tribe Ithomiini and the genus
\emph{Charaxes}, and for a decrease in net diversification rate in the
subtribe Danaina. Due to phylogenetic uncertainty, we did not obtain
strong support for other diversification splits in Nymphalidae. Our
BiSSE analysis corroborated other studies in that the trait 
``feeding on Solanaceae'', or a codistributed character, was important in the
diversification of Ithomiini butterflies. However, by applying MEDUSA we
found that a critical character in the radiation of the group might have
appeared after the shift from Apocynaceae to Solanaceae. We also found
that the trait ``feeding on Apocynaceae'' is not responsible for the
slowdown of diversification in Danaina. Ithomiini and Solanaceae
diversified near simultaneously, which is in agreement with the diffuse
co-speciation hypothesis \cite{nyman2012,janz2011}.

@@

\section*{Figure legends}
\begin{description}
 \item {\bf Figure 1. Results of the MEDUSA analysis run on the maximum
clade credibility tree from Wahlberg et al. (2009)} Rate shifts were estimated for the
following nodes (besides the background rate): 2) Limenitidinae +
Heliconiinae, 3) \emph{Ypthima}, 4) \emph{Charaxes}, 5) Ithomiini in
part, 6) Satyrini, 7) Coenonymphina 8) Phyciodina in part, 9) Danaini in
part, 10) \emph{Coenonympha}, 11) \emph{Caeruleuptychia} +
\emph{Magneuptychia}, 12) \emph{Callicore} + \emph{Diaethria}, 13)
Satyrina, 14) Mycalesina, 15) \emph{Pedaliodes}, 16) \emph{Dryas} +
\emph{Dryadula}, 17) \emph{Taenaris}, 18) Pseudergolinae. Circles on
nodes indicate the diversification shift number as found by MEDUSA.
Numbers next to circles indicat the posterior probability values for
such nodes.

>>>>

We contrasted these results with those obtained from analyzing the data in
the software BAMM \cite{rabosky2014}. We found 
that the best model (best configuration) estimated 2 diversification shifts in
our MCC tree (posterior probability = 0.79) for the Ithomiini
 and Satyrini. Bayes factor (BF) values strongly 
favored this model in comparison with a constant rate model (no diversification
shifts, BF $>$ 1,884).

@@

Our BiSSE analysis, extended to take into account missing taxa and
phylogenetic uncertainty, shows a significantly higher net
diversification rate for Ithomiini taxa, which can be attributed to the
trait ``feeding on Solanaceae hostplants'' (Fig. 4). This is in
agreement with previous findings using other statistical methods
\cite{fordyce2010}. Due to the fact that Ithomiini are virtually the only
nymphalids using Solanaceae as hostplants, it is possible that the trait
responsible for a higher diversification of Ithomiini might not be the
hostplant character. As noted by Maddison et al. (2007) \cite{maddison2007},
the responsible trait might be a codistributed character such as a trait
related to the ability to digest secondary metabolites.

@@

The increase in diversification inferred by MEDUSA occurred after the
probable shift from Apocynaceae to Solanaceae, as the Solanaceae feeders
in the subtribes Melinaeina and Mechanitina are not included in the
diversification shift (shift number 5 in Fig. 1). The apparent
conflicting results from MEDUSA and BiSSE can be explained by the low
species-richness of the subtribes Melinaeina and Mechanitina compared to
the other subtribes included in the shift (52 versus 272 species). It
can be that MEDUSA is more conservative than BiSSE and is not including
Melinaeina and Mechanitina in the shift due to low species numbers.

@@

Our MultiMEDUSA approach to perform analyses on the posterior
distribution of trees found strong support for an increase in net
diversification rate in the tribe Ithomiini and the genus
\emph{Charaxes}, and for a decrease in net diversification rate in the
subtribe Danaina. Due to phylogenetic uncertainty, we did not obtain
strong support for other diversification splits in Nymphalidae. Our
BiSSE analysis corroborated other studies in that the trait 
``feeding on Solanaceae'', or a codistributed character, was important in the
diversification of Ithomiini butterflies. However, by applying MEDUSA we
found that a critical character in the radiation of the group might have
appeared after the shift from Apocynaceae to Solanaceae. We also found
that the trait ``feeding on Apocynaceae'' is not responsible for the
slowdown of diversification in Danaina. Ithomiini and Solanaceae
diversified near simultaneously, which is in agreement with the diffuse
cospeciation hypothesis \cite{nyman2012,janz2011}.

@@

\section*{Figure legends}
\begin{description}
 \item {\bf Figure 1. Results of the MEDUSA analysis run on the maximum
clade credibility tree from Wahlberg et al. (2009)} Rate shifts were estimated for the
following nodes (besides the background rate): 2) Limenitidinae +
Heliconiinae, 3) \emph{Ypthima}, 4) \emph{Charaxes}, 5) Ithomiini in
part, 6) Satyrini, 7) Coenonymphina 8) Phyciodina in part, 9) Danaini in
part, 10) \emph{Coenonympha}, 11) \emph{Caeruleuptychia} +
\emph{Magneuptychia}, 12) \emph{Callicore} + \emph{Diaethria}, 13)
Satyrina, 14) Mycalesina, 15) \emph{Pedaliodes}, 16) \emph{Dryas} +
\emph{Dryadula}, 17) \emph{Taenaris}, 18) Pseudergolinae. Circles on
nodes indicate the diversification shift number as found by MEDUSA.
Numbers next to circles indicate the posterior probability values for
such nodes.
